### 2018-02-23 thesis/contents/discussions.tex
Updates from ShareLaTeX

\item Merged group tractography study of the trigeminal nerve and its primary sensory pathway as a novel use of Gaussian Process machine learning to the study of TN. We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in cross-sectional detail. Our report defines distinct bilateral changes in trigeminal neuralgia subjects, as well as changes between affected and unaffected sides.

@@

The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus signified that for the first time we could use software methods to segment the pontine trigeminal anatomy in a reliable manner. Prior to this, the ROI identification was done visually, and sometimes with simple heuristics methods based on distance from surrounding landmarks. Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be always assessed.

The brainstem CN V nerve in some subjects could be seen to bend caudally into the medulla. Althought not reported in Study IV, sometime we could delineate what appeared to be the spinal trigeminal decussation at the level of the medulla. The fibers were sparse, and not consistently reproducible across our subjects. Moreover, the mesencephalic trigeminal nucleus could be delineated in some subjects as well. Merge tractography demonstrated that these anatomical features have low reproducibility, and therefore they were not included as part of the study. In theory, with enough resolution, these smaller structures can be imaged. It would also be interesting to attempt to delineate the motor branch of the trigeminal nerve.

Diffusion tensor imaging revolutionarized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter is a limitation of diffusion tractography. The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging. 
Techniques to improve visualization of white matter fibers consist in probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map. This kind of image-based probabilistic tractography was useful for whole-brain white matter studies. However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways. Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition. The acquisition time of these sequences was impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development. SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions. This meant that these methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.

These algorithmic developments brought imaging of crossing-fibers using tractography to the realm of possibility. We explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST. The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images. 2) it courses through single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.

The segmentation of the subregions of the trigeminal pathway has allowed for the possibility of measuring the respective diffusivities. Moreover, this method can extend to any scalar images that can be registered to the DWI space. These include specific tissue-metrics such as myelin, G-ratio, and PET images. This methodology is therefore highly applicable and not specific to the study of TN. We divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern. The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure. In theory the number of divisions could be infinitely large, given that a centroid curve was identified. In practice the cross-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered. Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF. White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chaism. A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research. Turns out that the merged tractography strategy simplified this task by allowed a single curve to be defined manually in the template tractography space. This greatly reduced the amount of labour for researchers, as centroids no longer need to be defined repeatedly for each subject. Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.

@@

We applied machine learning to the analysis of TN in Study IV in order to avoid our own confirmation bias. The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method. We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret. This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics. Eventually it was decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which contradicts frequentist assumptions. The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency. 2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive. More specifically, the streamline measurements resemble time-series data, in that it exhibits positional dependency between measurements. Unlike a time-series however, the streamline measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency, as suppose to uni-directional in time. Posing the study as a supervised machine learning problem, the feature vector would be a N dimension vector representing N measures at each division, and a classification label. SVM would assume the features to be points in a N-dimensional space, and each measurement would assume independence. This makes the result of the SVM output difficult to interpret. The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output. Therefore the GP model is able to take the positional dependency into account, and the result output is also much more intuitive. Moreover, changing the GP classification model into a regression one is trivial. This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a patient's CN V, what is the probability that is resembles a pathological population, what about a healthy population? It is possible to answer these questions trivially with the GP model using regression. The full implication is outside of the scope of this thesis, but its implementation is has already been seeded within.

@@

The biggest value of tractography to neurosugery is to assist in pre-operative planning. Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D. We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}.
Another contribution is to visualize tissue changes from surgical interventions that can otherwise be invisible. There are cases where post-operative GammaKnife ablation target on the trigeminal nerve did not result in dicernable gadonenium enhancement in standard MR, but the nerve FA changes can be seen in the delineated tractography. Therefore tractography can provide both spatial, as well as tissue-based insights to the surgeons to aid in assessment.
Tractography and 3D anatomical model ultimately are standard 3D geometries in computer graphics. Therefore advances in 3D printing can be easily adopted in neurosurgery to allow surgeons to physically examine tractography reconstructions. On the otherhand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making. Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.

@@

From my observations across the studies in TN, I postulate a resonance hypothesis of TN. I am aware that there lacks substantial physiological measurements for these claims, nevertheless I feel it is important to state them here as a possible direction of investigation for the future: In Gamma-Knife radiosurgery treatment for TN, the target is on the cistern CN V segment, and situates proximally to the CN V REZ. Paradoxically, the GK treatment can be seen as a willful disruption of the CN V myelination. The question is, why would additional GK lesions on the CN V myelin effectively decrease TN pain. Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain. The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved. Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure. The force exerted onto the CN V from its neighbouring vasculature is therefore periodic. It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition". This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain. Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain. This model also explains that GK intervention disrupts the chance of resonance condition, thereby reducing TN pain. The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns. MVD, in this case can be explained as the physical disruption of the condition.

>>>>

\item Merged group tractography study of the trigeminal nerve and its primary sensory pathway as a novel use of Gaussian Process machine learning to the study of TN. We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (thalamocortical, S1) white patter pathways in sub-regional detail. Our report defines distinct bilateral changes in trigeminal neuralgia subjects, as well as changes between affected and unaffected sides.

@@

The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus signified that for the first time we could use software methods to segment the pontine trigeminal anatomy in a reliable manner. Prior to this, the ROI identification was done visually, and sometimes with simple heuristics methods based on distance from surrounding landmarks. Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be always assessed. Improvements in tractography of the brainstem trigeminal nerve allowed the shift towards more algorithmic driven and automated segmentation and ROI placement.

Diffusion tensor imaging revolutionized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter is a limitation of diffusion tractography. The brain white matter consists of complex crossing fiber crossings, and the inability for DTI to reliably image these structure therefore posed a fundamental challenge for neuroimaging.
Techniques to improve visualization of white matter fibers consist in probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map. This kind of image-based probabilistic tractography was useful for whole-brain white matter studies. However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways. Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition. The acquisition time of these sequences was impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research.
The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development. SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions. This meant that these methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.

These algorithmic developments brought imaging of crossing-fibers using tractography to the realm of possibility. We explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST. The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images. 2) it courses through simple (cistern) and complex (pons) fiber environments; 3) it is a prominent cranial nerve, and therefore a good benchmark for imaging small anatomy.

We divided the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern. The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure. In theory the number of divisions could be infinitely large, given that a centroid curve was identified. In practice the sub-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered. Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF. White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chiasm. A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research. 
Turns out that the merged tractography strategy simplified this task by allowed a single curve to be defined manually in the template tractography space. This greatly reduced the amount of labour for researchers, as centroids no longer need to be defined repeatedly for each subject. Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.

The segmentation of the subregions of the trigeminal pathway has allowed for the possibility of measuring the respective diffusivities. Moreover, this method can extend to any scalar images that can be registered to the DWI space. These may include specific tissue-metrics such as myelin, G-ratio, and PET images. This methodology is therefore highly applicable and not specific to the study of TN.

@@

We applied machine learning to the analysis of TN in Study IV in order to avoid our own confirmation bias. The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method. We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret. This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics. Eventually it was decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which may not fit with frequentist assumptions. The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency. 2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive. More specifically, the streamline measurements resemble time-series data, in that it exhibits positional dependency between measurements. Unlike a time-series however, the streamline measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency, as suppose to uni-directional in time. Posing the study as a supervised machine learning problem, the feature vector would be a N dimension vector representing N measures at each division, and a classification label. SVM would assume the features to be points in a N-dimensional space, and each measurement would assume independence. This makes the result of the SVM output difficult to interpret. The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output. Therefore the GP model is able to take the positional dependency into account, and the result output is also much more intuitive. Moreover, changing the GP classification model into a regression one is trivial. This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a patient's CN V, what is the probability that is resembles a pathological population, what about a healthy population? It is possible to answer these questions trivially with the GP model using regression. The full implication is outside of the scope of this thesis, but its implementation is has already been seeded within.

@@

Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D. We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}.
Another contribution is to visualize tissue changes from surgical interventions that can otherwise be invisible. There are cases where post-operative GammaKnife ablation target on the trigeminal nerve did not result in dicernable gadonenium enhancement in standard MR, but the nerve FA changes can be seen in the delineated tractography. Therefore tractography can provide both spatial, as well as tissue-based insights to the surgeons to aid in assessment.
Tractography and 3D anatomical model ultimately are standard 3D geometries in computer graphics. Therefore advances in 3D printing can be easily adopted in neurosurgery to allow surgeons to physically examine tractography reconstructions. On the otherhand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making. Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.

@@

From my observations across the studies in TN, I postulate a resonance hypothesis of TN. I am aware that there lacks substantial physiological measurements for these claims, nevertheless I feel it is important to state them here as a possible direction of investigation for the future: In surgical interventions for TN, the target is on the cistern CN V segment, and situates close to the CN V REZ. Paradoxically, the intervention can be seen as a willful disruption of the CN V myelination. The question is, why would additional disruptions on the CN V myelin effectively decrease TN pain. Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain. The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved. Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure. The force exerted onto the CN V from its neighbouring vasculature is therefore periodic. It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition". This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain. Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain. This model also explains that surgical intervention disrupts the chance of resonance condition, thereby reducing TN pain. The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns. MVD, in this case can be explained as the physical disruption of the condition.

### 2018-02-23 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

\subsubsection{Along-the-tract Analysis}
For the purpose of quantifying the trigeminal pathways, a centroid was defined for the merged tractography of each pathway (Figure \ref{fig:GPfigure-method-centroid}). The centroids were defined as the average curve that best fit the tract bundles, and were in template space. CN V centroids were manually defined using 3 control points. A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}. The spline was then resampled into 30 points. TPT and S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30. Nearest-neighbor point assignment was used to cluster  the points of the merged tract model to the centroid control points, to obtain the corresponding 30 sections along each centroid. Diffusion metrics from the tractography points were averaged for each of the sections for each subject.

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-path-analysis.pdf}
\caption{Diagram for the along-the-tract data processing.}
\label{fig:GPfigure-method-centroid}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-GP-training.pdf}
\caption{Diagram for generating the best performing classifier for a particular window size. The data for each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position. Finally the window position with the best accuracy was selected. The same process is repeated on every diffusivity metric, and for all window sizes ranging from 5 to 30. }
\label{fig:GPfigure-gp-train}
\end{figure}

@@

\subsubsection{Diffusivity Differences}
\paragraph{CN V}
The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves.Both sides were distinguished from controls with classifiers that achieved around 80\% accuracy (Figure \ref{fig:GPfigure4}). 
Classifiers on the right side had accuracies ranging from 76\% (GFA) to 86\% (FA). The left classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA). The right side showed more overlap of segment ranges between positions 14--24, in the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4 in the brainstem. Overall, the right side classifiers extend further into the distal space, into the cistern and even the ganglion, while the left side extended more into the brainstem. FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cistern).
The FA classifier achieved 75\% accuracy in differentiating affected versus unaffected nerves, where its segments ranged from 11 to 19, which was the cisternal segment. The GFA classifier achieved 70\% accuracy in region 9--12. Demonstrating that affected-unaffected diffusivity differences were found in the cistern/REZ (Table \ref{table:CN V}).

@@

\begin{table}[ht]
\centering
\csvautotabular{cnv_table.csv}
\caption{CN V GP classifier performance data}
\caption*{List of the best accuracy for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
\label{table:CN V}
\end{table}

\paragraph{Trigeminopontothalamic tract}
Classifiers could distinguish bilateral TPT pathways (affected and unaffected) from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}). The right FA classifier achieved the greatest accuracy at 85\%, and right RD classifier was least accuracy at 83\%. The left FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuray at 71\%. The classifiers covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage. The left-versus-right classifiers achieved maximum accuracy of 98\% (FA) and minimum of 88\% (MD). The FA classifier covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:tpt}).

@@

\begin{table}[ht]
\centering
\csvautotabular{tpt_table.csv}
\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
\label{table:tpt}
\end{table}

\paragraph{S1}
Classifiers distinguished bilateral S1 face region pathways from controls with about 85\% accuracy. The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at position 8--26 (Figure \ref{fig:GPfigure5}). Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifiers ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:s1}). Classifiers of both sides covered the the branching region between the thalamus and cortex. Left-versus-right classifiers showed maximum accuracy of 78\% (RD), and minimum of 76\% (FA). The left-versus-right classifiers overlap in the regions 5--10, at the segment where the pathway emerges from the thalamus.

@@

\begin{table}[ht]
\centering
\csvautotabular{s1_table.csv}
\caption{S1 GP classifiers performance data}
\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
\label{table:s1}
\end{table}

>>>>

\subsubsection{Along-the-tract Analysis}
For the purpose of quantifying the trigeminal pathways, a centroid was defined for the merged tractography of each pathway (Figure \ref{fig:GPfigure-method-centroid}). The centroids were defined as the average curve that best fit the tract bundles, and were in template space. CN V centroids were manually defined using 3 control points. A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}. The spline was then resampled into 30 subpoints. TPT and S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30. Nearest-neighbor point assignment was used to cluster the points of the merged tract model to the centroid subpoints, to obtain the corresponding 30 divisions along each centroid. Diffusion metrics from the tractography points were averaged for each of the divisions for each subject.

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-path-analysis.pdf}
\caption{Diagram for the along-the-tract data processing.
Process starts with 2 specific goals: 1) Derive the spatial structure that best represent the merged streamlines, 2) Deriving sub-regional tissue information from these. A) First the best-fit curve is obtained to represent the collective streamlines, which is defined as the centroid. B) In areas where this can be automated, Quickbundles was used. For the CN V, the centroid was defined manually given the unpredictability of the lengths of streamlines. C) Representative drawing of the centroid onto the streamline points, representing the collective spatial information. The centroid is resampled to 30 subpoints. D) To derive tissue information, a nearest neighbor algorithm was used to assign streamline points to form 30 divisions. E) The average diffusivity metrics was pooled for each of the 36 subjects at each division. The colours represent the spatial position of each of the 30 divisions, and the streamline points that were assigned to each particular division. The data was then used for machine learning pipeline (see Figure \ref{fig:GPfigure-gp-train}).
}
\label{fig:GPfigure-method-centroid}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{thesis/images/tn-gp-sagit/figure-GP-training.pdf}
\caption{Diagram for generating the best performing classifier for a particular window size. The data was derived from Figure \ref{fig:GPfigure-method-centroid}, each window was used to generate 10 GP models with 10-fold cross-validation, an average accuracy was determined at each window position. Finally the window position with the best accuracy was selected. The same process is repeated on every diffusivity metric, and for all window sizes ranging from 5 to 30. }
\label{fig:GPfigure-gp-train}
\end{figure}

@@

\subsubsection{Diffusivity Differences}
\paragraph{CN V}
The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves.Both sides were distinguished from controls with classifiers that achieved around 80\% accuracy (Figure \ref{fig:GPfigure4}). 
Classifiers on the right side had accuracies ranging from 76\% (GFA) to 86\% (FA). The left classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA). The right side showed more overlap of segment ranges between positions 14--24, in the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4 in the brainstem. Overall, the right side classifiers extend further into the distal space, into the cistern and even the ganglion, while the left side extended more into the brainstem. FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cistern).
The FA classifier achieved 75\% accuracy in differentiating affected versus unaffected nerves, where its segments ranged from 11 to 19, which was the cisternal segment. The GFA classifier achieved 70\% accuracy in region 9--12. Demonstrating that affected-unaffected diffusivity differences were found in the cistern/REZ (Table \ref{table:GP}).

@@

\paragraph{Trigeminopontothalamic tract}
Classifiers could distinguish bilateral TPT pathways (affected and unaffected) from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}). The right FA classifier achieved the greatest accuracy at 85\%, and right RD classifier was least accuracy at 83\%. The left FA classifier showed greatest accuracy at 89\% and left RD classifier showed the least accuray at 71\%. The classifiers covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage. The left-versus-right classifiers achieved maximum accuracy of 98\% (FA) and minimum of 88\% (MD). The FA classifier covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:GP}).

@@

\paragraph{S1}
Classifiers distinguished bilateral S1 face region pathways from controls with about 85\% accuracy. The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at position 8--26 (Figure \ref{fig:GPfigure5}). Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifiers ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:GP}). Classifiers of both sides covered the the branching region between the thalamus and cortex. Left-versus-right classifiers showed maximum accuracy of 78\% (RD), and minimum of 76\% (FA). The left-versus-right classifiers overlap in the regions 5--10, at the segment where the pathway emerges from the thalamus.

@@

\begin{table}[ht]
\centering
\csvautotabular{tn_sagit_table.csv}
\caption{List of the best GP models for different diffusivity meatures and window segments. Only models with at least 70\% accuracy were accepted. Precision and f1 scores are also provided for reference}
\label{table:GP}
\end{table}
